Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2019-04-08 WholePlotTrajectories.Rmd
review Bruno

---
title: "Post-Disturbance Tree Community Trajectories in a Neotropical Forest"
author:
  - name: Ariane MIRABEL
    authsuperscript: 1*
  - name: Bruno Herault
    authsuperscript: 2
  - name: Eric Marcon
    authsuperscript: 1
affiliation:
  - affsuperscript: 1
    dptuniv: UMR EcoFoG, AgroParistech, CNRS, Cirad, INRA, Université des Antilles, Université de Guyane.
    address: >
      Campus Agronomique,
      97310 Kourou,
      France.
  - affsuperscript: 2
    dptuniv: INPHB, Institut National Polytechnique Félix Houphoüet-Boigny
    address: >
      Yamoussoukro,
      Ivory Coast.
corrauthor:
    email: ariane.mirabel@ecofog.gf
    url: https://github.com/ArianeMirabel
abstract: 
 Anticipate the impact of the global change context on tropical forests is now urgent.
 This calls to highlight the taxonomic and functional facets of tree community post-disturbance trajectories and the underlying processes. 
 It also calls to clarify the link between biodiversity and disturbance, specifically the scope of the Intermediate Disturbance Hypothesis (IDH) that is debated in tropical forests.
 In this study, we analyzed the tree community trajectories following a disturbance gradient in a Neotropical forest over 30 years. 
 We considered community taxonomic and functional diversity, composition and redundancy trajectories. We based on the annual botanical inventories of 75 ha of a Neotropical forest and on large trait datasets comprising 7 leaf, stem and life-history traits. 
 We highlighted a decoupling between taxonomic trajectories, that differed among communities, and functional trajectories, remained similar and convergent. We explained this decoupling by the variations in community functional redundancy that mitigated the functional impact of disturbance. Post-disturbance trajectories clarified the scope of the Intermediate Disturbance Hypothesis that was validated regarding taxonomic diversity, with an intensity threshold of 20-25%, but not regarding functional diversity.The IDH applied around 20 years after disturbance when the recruitment of pioneers and late-successionals were balanced. 
 Although consistent, the recovery of community composition, diversity and redundancy remained unachieved after 30 years. This acknowledged the need of decades-long cycles without disturbance to ensure a complete recovery, and questioned tropical forest community resilience after repeated disturbances.
keywords: [Community Ecology, Disturbance Trajectories, Intermediate Disturbance Hypothesis, Mid-term Resilience, Neotropical Forests, Taxonomic and Functional Biodiversity]
preamble: >
  \usepackage{textcomp}
  \DeclareUnicodeCharacter{B0}{\textdegree}
  \usepackage{tabu}
  \usepackage{caption}
  \captionsetup{justification = justified}
  \renewenvironment{table}{\begin{table*}}{\end{table*}\ignorespacesafterend}
  \hyphenation{bio-di-ver-si-ty sap-lings re-or-gan-i-za-tion post-dis-tur-bance dis-tur-bance}
bibliography: references.bib
lang: english
keywordlabel: Keywords
corrauthorlabel: Corresponding author
output:
  bookdown::pdf_book:
    base_format: EcoFoG::article
    keep_tex: TRUE
  bookdown::word_document2: default
  bookdown::gitbook: default
---

@@

The experiment is a network of twelve 6.25 ha plots (Table \@ref(tab:Tab1)) that underwent three disturbance treatments in 1987 according to a randomized plot design [@Gourlet-Fleury2004].  
The experiment comprised three replicates of three sylvicultural treatments (hereafter plots T1, T2 and T3) and three control plots (T0).
All treatments T1, T2 and T3 comprised the logging of 10 trees/ha with 50 cm minimum DBH that belonged to a set of 58 commercial species [@Gourlet-Fleury2004]. 
Treatment T2 additionally comprised a thinning treatment by poison-girdling of non-commercial, randomly selected species with an average 30 trees/ha with 40 cm minimum DBH.
Treatment T3 additionally comprised the logging of 15 trees /ha with 40 cm minimum DBH and the poison-gidling of 20 trees/ha with a 50 cm minimum DBH, all belonging to non-commercial species. 
Considering the silvicultural treatments and the following damage, disturbance intensity was measured as the percentage of aboveground biomass (%AGB) lost between the first inventory in 1984 and five years after disturbance [@Piponiot2016] estimated with the BIOMASS R package [@Biomass2018].
The three treatments constituted a disturbance intensity gradient with increasing of above-ground biomass (AGB) lost and surface disturbed.

```{r Tab1, echo=FALSE,as.is=T}
Intervention <- data.frame(
  Treatment=c("Control", "T1, low", "T2, intermediate", "T3, high"),
  Timber=c("-", 
           "DBH $\\geq$ 50 cm, commercial species, $\\approx$ 10   $trees.ha^{-1}$", 
           "DBH $\\geq$ 50 cm, commercial species, $\\approx$ 10  $trees.ha^{-1}$",
           "DBH $\\geq$ 50 cm, commercial species, $\\approx$ 10  $trees.ha^{-1}$"),
  Thinning=c("-", 
             "-", 
             "DBH $\\geq$ 40 cm, non-valuable species, $\\approx$ 30   $trees.ha^{-1}$", 
             "DBH $\\geq$ 50 cm, non-valuable species, $\\approx$ 15  $trees.ha^{-1}$  "),
  Fuelwood=c("-", 
             "-", 
             "-", 
             "40 cm $\\leq$ DBH $\\leq$ 50 cm, non-valuable species,\\ $\\approx$ 15 $trees.ha^{-1}$"),
  AGBLost=c(0, "$[12-33]$", "$[33-56]$", "$[35-56]$")
)
names(Intervention)[5] <- "\\%AGB lost"
knitr::kable(Intervention, caption="Intervention table, summary of the disturbance intensity for the 4 plot treatments in Paracou.", longtable = FALSE, booktabs = TRUE, escape = FALSE,format="latex") %>%
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = TRUE)
```

@@

Six functional traits representing leaf economics (leaf thickness, toughness, total chlorophyll content and specific leaf area) and stem economics (wood specific gravity and bark thickness), and life-history traits (maximum specific height and seed mass) came from the BRIDGE project [^1]. 
Trait values were assessed from a selection of individuals located in nine permanent plots in French Guiana, including two in Paracou, and comprised 294 species pertaining to 157 genera. 
Missing trait values (10%) were filled using multivariate imputation by chained equation [@Mice2011].
Imputations were restricted within genus or family when samples were too scarce, in order to account for the phylogenetic signal. 
Whenever a species was not in the dataset, it was attributed a set of trait values randomly sampled among species of the next higher taxonomic level (same genus or family). 
As seed mass information was classified into classes, no data filling process was applied and analyses were restricted to the 414 botanical species recorded.

@@

[^1]: http://www.ecofog.gf/Bridge/

@@

Because of the variability in the precision of botanical identification efforts, we were constrained to conduct the taxonomic composition and diversity analysis at the genus level. 
Taxonomic and functional trajectories of community composition were followed in a two-dimensional NMDS ordination plane.
Two NMDS using abundance-based (Bray-Curtis) dissimilarity measures were conducted to map either taxonomic or functional composition, the latter based on the seven leaf, stem and life history traits (without seed mass classes).
Trajectories along time were reported through the Euclidean distance between the target inventories and the reference inventories in 1989, *i.e.* 2 years after disturbance. 
Univariate trajectories of the leaf, stem and life-history traits were also visualized with the community weighted means (CWM) [@Diaz2007]. 
Species seed mass were given in 5 mass classes, and seed mass trajectories were reported as the proportion of each class in the inventories (Supplementary materials).

The taxonomic diversity was reported through species richness and the Hill number translation of the Simpson index [@Hill1973].
These metrics allowed assessing the taxonomic richness as well as evenness, through the comparison between these two metrics.: results will be discussed directly in terms of taxonomic richness and evenness.
Both indices belong to the set of HCDT or generalized entropy, respectively corresponding to the 0 and 2 order of diversity (q), recommended for diversity studies [@Marcon2015b]. 
The functional diversity was reported using the functional richness and functional evenness, *i.e* the Rao index of quadratic entropy.
The Rao index combines species abundance distribution, and the average pairwise functional dissimilarity between species computed by the gower distance.

@@

From 1989 (2 years after disturbance) to 2015 (28 years after disturbance), 828 388 individual trees and 591 botanical species pertaining to 223 genera and 64 families were recorded.

@@

par(mfrow=c(2,2),mar=c(2,2.5,2.5,1),oma=c(1,2,1,4),no.readonly = T)
TaxoCompo(MatrepTaxo)
mtext("Taxonomic composition",side=3,line=1.8,cex=1.05)
FunCompo(MatrepFun)
mtext("Functional composition",side=3,line=1.8,cex=1.05)
EuclidDist(TaxoEuclid)
mtext("Euclidean distance\nfrom 1985 inventory",side=2,padj=0,line=2,cex=0.8)
mtext("(c)",side=3,adj=0,line=0.5)
EuclidDist(FunEuclid)
mtext("(d)",side=3,adj=0,line=0.5)
mtext("Years since disturbance",side=1,line=2.2,adj=1,cex=0.8)
legend("right",inset=-0.4,xpd=NA,legend=c("Control","Low","Interm.","High"),col=ColorsTr,lwd=2.5,bty="n",title="Treatment", cex=0.85)

@@

```{r DivTaxo, out.width = '100%', echo=FALSE, fig.cap="Trajectories over 30 years of the difference with the 1989 inventory (2 years after disturbance) of community taxonomic richness (ref:Bolda), Simpson diversity (ref:Boldb), functional richness (ref:Boldc), and Rao diversity (ref:Boldd). Shaded areas are the credibility intervals ", fig.env="figure*"}
load("DB/ReplacementTraj_ForGraphs")
load("DB/FunctionalTraj_ForGraphs")
load("DB/FunctionalRichnessTraj_ForGraphs")

@@

@@

Before disturbance, the different communities had different taxonomic compositions, visualized by their distinct starting points on the NMDS axis 2.
These initial differences were maintained all along the 30 years following disturbance, with the disturbance leading a displacement on the NMDS axis 1.
The taxonomic compositional changes following disturbance were similar among communities, which may correspond to the recruitment of a group of pioneers common to all communities whenever their initial differences and the intensity of disturbance [@Denslow2000;@Bongers2009].
Taxonomic trajectories continued with a recovery of the initial composition which, althought not fully achieved after 30 years, suggested the resilience of the taxonomic composition resilience and the maintenance of the initial differences among communities [@Folke2006].
The  return towards initial taxonomic composition showed that species not belonging to the pre-disturbance community were rarely recruited, probably because of the common dispersal limitations among tropical tree species [@Svenning2005].

@@

Community taxonomic richness and evenness trajectories were determined by the disturbance intensity, ranging from a limited and temporary impact to significant and persistant alterations of community diversity.
The taxonomic trajectories markedly changed above an intensity threshold at which the taxonomic richness was maximized and the taxonomic evenness remained resilient. 
The disturbance intensity impacted the balance in the community between pre-disturbance surviving trees, and trees recruited afterward. 
On the one hand, below a 20%-25% AGB loss, the trees surviving after disturbance remained numerous enough to maintain the high taxonomic richness of the pre-disturbance community [@Bongers2009].
The recruitment of pioneers, infrequent or absent before disturbance, then increased the taxonomic richness all the more so that the disturbance was intense [@Martin2015; @Chaudhary2016].
As these pioneers became more dominant they balanced the usual hyper-dominance of tropical forests, hence temporarily increasing the taxonomic evenness [@Baraloto2012a].
On the other hand, above the intensity threshold, the taxonomic richness of surviving trees was too low to be offset by the recruitment of pioneers. 
In the Guiana shield indeed, the pool of true pioneers specifically recruited after disturbance is restricted to a few common genera (e.g. *Cecropia* spp., *Vismia* spp.) [@Guitet2018].
The taxonomic richness then decreased following intense disturbance, all the more so that the disturbance intensity was high [@Molino2001]. 
At the same time, pioneers became persistently dominant and prevented a return towards the initial evenness and the recovery of hyper-dominant shade-tolerant species.

@@

The decoupling between taxonomic and functional trajectories was explained by a decrease in the functional redundancy within the pre-disturbance functional space, due to the loss of species following disturbance.
The recruitment of pioneers, functionally different from the pre-disturbance functional composition, did not compensate the decrease of functional redundancy in the first place.
Progressively though, the functional redundancy was restored through the replacement of the first established species by more competitive, long-lived pioneers or late-successional species that were functionally closer to the pre-disturbance community.
This replacement followed the lottery recruitment rules, implying an easy recruitment for the first species but becoming increasingly hampered by the emergence of interspecific competition [@Busing2002]. 
The recovery of the functional redundancy then relied upon the random process of species recruitment and was increasingly slow and difficult to anticipate [@Elmqvist2003;@Diaz2005].

@@

Community post-disturbance trajectories in compositon and diversity were shaped by the recruitment of a determined pool of pioneers identical among communities and disturbance intensity.
Nevertheless, taxonomic and functional composition trajectories appeared decoupled.
While functional trajectories remained similar in the functional space and converged towards the recovery of a comparable initial state, taxonomic trajectories showed initial differences in community composition that were maintained along time.
This decoupling was explained by a decrease of the functional redundancy that mitigated the functional impact of disturbance.
Community functional and taxonomic diversity response to disturbance was contrasted as well.
While the functional trajectories remained similar whenever the disturbance intensity, taxonomic trajectories were markedly different above a threshold of 20-25% AGB removed for which taxonomic richness was maximized and taxonomic eveness remained resilient.
The Intermediate Disturbance Hypothesis applied to the linkage of disturbance with the taxonomic diversity, but not with the functional diversity.
Taxonomic diversity was maximized following a disturbance of intermediate intensity, that was the threshold of 25% AGB removed, and at an intermediate time after disturbance, that was around 25 years after disturbance when the recruitment of early- and late-successional species were balanced.
Whenever the disturbance intensity, community resilience (in terms of recovery of the pre-disturbance state) was tangible but required several decades and relied upon the random lottery recruitment of rare species. 
Although resilient, the functional redundancy was lower for more than 30 years after disturbance which cautioned against the risks of infrequent species loss and the persistence of disturbance-specific communities [@Herault2018].

>>>>

---
title: "Post-Disturbance Tree Community Trajectories in a Neotropical Forest"
author:
  - name: Ariane MIRABEL
    authsuperscript: 1*
  - name: Bruno Herault
    authsuperscript: 2
  - name: Eric Marcon
    authsuperscript: 1
affiliation:
  - affsuperscript: 1
    dptuniv: UMR EcoFoG, AgroParistech, CNRS, Cirad, INRA, Université des Antilles, Université de Guyane.
    address: >
      Campus Agronomique,
      97310 Kourou,
      France.
  - affsuperscript: 2
    dptuniv: INPHB, Institut National Polytechnique Félix Houphoüet-Boigny
    address: >
      Yamoussoukro,
      Ivory Coast.
corrauthor:
    email: ariane.mirabel@gmail.com
    url: https://github.com/ArianeMirabel
abstract: 
 Anticipate the impact of the global change context on tropical forests is now urgent.
 This calls to highlight the taxonomic and functional facets of tree community post-disturbance trajectories and the underlying processes. 
 It also calls to clarify the link between biodiversity and disturbance, specifically the scope of the Intermediate Disturbance Hypothesis (IDH) that is debated in tropical forests.
 In this study, we analyzed the tree community trajectories following a disturbance gradient in a Neotropical forest over 30 years. 
 We considered community taxonomic and functional diversity, composition and redundancy trajectories. We based on the annual botanical inventories of 75 ha of a Neotropical forest and on large trait datasets comprising 7 leaf, stem and life-history traits. 
 We highlighted a decoupling between taxonomic trajectories, that differed among communities, and functional trajectories, remained similar and convergent. We explained this decoupling by the variations in community functional redundancy that mitigated the functional impact of disturbance. Post-disturbance trajectories clarified the scope of the Intermediate Disturbance Hypothesis that was validated regarding taxonomic diversity, with an intensity threshold of 20-25%, but not regarding functional diversity.The IDH applied around 20 years after disturbance when the recruitment of pioneers and late-successionals were balanced. 
 Although consistent, the recovery of community composition, diversity and redundancy remained unachieved after 30 years. This acknowledged the need of decades-long cycles without disturbance to ensure a complete recovery, and questioned tropical forest community resilience after repeated disturbances.
keywords: [Community Ecology, Disturbance Trajectories, Intermediate Disturbance Hypothesis, Mid-term Resilience, Neotropical Forests, Taxonomic and Functional Biodiversity]
preamble: >
  \usepackage{textcomp}
  \DeclareUnicodeCharacter{B0}{\textdegree}
  \usepackage{tabu}
  \usepackage{caption}
  \captionsetup{justification = justified}
  \renewenvironment{table}{\begin{table*}}{\end{table*}\ignorespacesafterend}
  \hyphenation{bio-di-ver-si-ty sap-lings re-or-gan-i-za-tion post-dis-tur-bance dis-tur-bance}
bibliography: references.bib
lang: english
keywordlabel: Keywords
corrauthorlabel: Corresponding author
output:
  bookdown::pdf_book:
    base_format: EcoFoG::article
    keep_tex: TRUE
  bookdown::word_document2: default
  bookdown::gitbook: default
---

@@

The experiment is a network of twelve 6.25 ha plots (Table \@ref(tab:Tab1)) that underwent three disturbance treatments in 1987 according to a randomized plot design [@Gourlet-Fleury2004].  
The experiment comprised three replicates of three sylvicultural treatments (hereafter plots T1, T2 and T3) and three control plots (T0).
All treatments T1, T2 and T3 comprised the logging of 10 trees/ha with 50 cm minimum DBH that belonged to a set of 58 commercial species [@Gourlet-Fleury2004]. 
Treatment T2 additionally comprised a thinning treatment by poison-girdling of non-commercial, randomly selected species with an average 30 trees/ha with 40 cm minimum DBH.
Treatment T3 additionally comprised the logging of 15 trees /ha with 40 cm minimum DBH and the poison-gidling of 20 trees/ha with a 50 cm minimum DBH, all belonging to non-commercial species. 
Considering the silvicultural treatments and the following damage, disturbance intensity was measured as the percentage of aboveground biomass (%AGB) lost between the first inventory in 1984 and five years after disturbance [@Piponiot2016] estimated with the BIOMASS R package [@Rejou2017].
The three treatments constituted a disturbance intensity gradient with increasing of above-ground biomass (AGB) lost and surface disturbed.

```{r Tab1, echo=FALSE,as.is=T}
Intervention <- data.frame(
  Treatment=c("Control", "T1, low", "T2, intermediate", "T3, high"),
  Timber=c("-", 
           "DBH $\\geq$ 50 cm, commercial species, $\\approx$ 10   $trees.ha^{-1}$", 
           "DBH $\\geq$ 50 cm, commercial species, $\\approx$ 10  $trees.ha^{-1}$",
           "DBH $\\geq$ 50 cm, commercial species, $\\approx$ 10  $trees.ha^{-1}$"),
  Thinning=c("-", 
             "-", 
             "DBH $\\geq$ 40 cm, non-valuable species, $\\approx$ 30   $trees.ha^{-1}$", 
             "DBH $\\geq$ 50 cm, non-valuable species, $\\approx$ 15  $trees.ha^{-1}$  "),
  Fuelwood=c("-", 
             "-", 
             "-", 
             "40 cm $\\leq$ DBH $\\leq$ 50 cm, non-valuable species,\\ $\\approx$ 15 $trees.ha^{-1}$"),
  AGBLost=c(0, "$[12-33]$", "$[33-56]$", "$[35-56]$")
)
names(Intervention)[5] <- "\\%AGB lost"
knitr::kable(Intervention, caption="Intervention table, summary of the disturbance intensity for the 4 plot treatments in Paracou. Treatment intensities are defined by the minimum logging DBH (Diameter at Breast Height), the type of logged species (commercial or not), the density of logged trees, and the total AGB (Above Ground Biomass) lost after treatment.", longtable = FALSE, booktabs = TRUE, escape = FALSE,format="latex") %>%
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = TRUE)
```

@@

Six functional traits representing leaf economics (leaf thickness, toughness, total chlorophyll content and specific leaf area) and stem economics (wood specific gravity and bark thickness) came from the BRIDGE project [^1]. 
Trait values were assessed from a selection of individuals located in nine permanent plots in French Guiana, including two in Paracou, and comprised 294 species pertaining to 157 genera. 
Whenever a species was in the dataset but missed some trait values (10% of the species), missing values were filled using multivariate imputation by chained equation [@Mice2011].
To account for the phylogenetic signal in the filling process, imputations based on samples from the same genus or from the same family. 
Whenever a species was not in the dataset, it was attributed a set of trait values randomly sampled among species of the next higher taxonomic level (same genus or family). 
Two life-history traits (maximum specific height and seed mass) came from the Mariwenn database [^2]. 
The database compiles information from a vast literature on the flora of French Guiana [@Ollivier2007] and comprises 362 species pertaining to 188 genera.
As seed mass information was classified into classes, no data filling process was applied and analyses were restricted to the botanical species recorded.

@@

[^1]: http://www.ecofog.gf/Bridge/ 
[^2]: https://www.ecofog.gf/mariwenn/

@@

Because of the variability in the precision of botanical identification efforts, we were constrained to conduct the taxonomic composition and diversity analysis at the genus level. 
Taxonomic and functional trajectories of community composition were followed in a two-dimensional NMDS ordination plane.
Two NMDS using abundance-based (Bray-Curtis) dissimilarity measures were conducted to map either taxonomic or functional composition, the latter based on the seven leaf, stem and life history traits (without seed mass classes).
Trajectories along time were reported through the Euclidean distance between the target inventories and the reference pre-disturbance inventories of 1984.
Univariate trajectories of the leaf, stem and life-history traits were also visualized with the community weighted means (CWM) [@Diaz2007]. 
Species seed mass were given in 5 mass classes, and seed mass trajectories were reported as the proportion of each class in the inventories (Supplementary materials).

The taxonomic and functional trajectories were analysed from the reference pre-disturbance inventories of 1984.
The taxonomic diversity was reported through species richness and the Hill number translation of the Simpson index [@Hill1973].
These metrics allowed assessing the taxonomic richness as well as evenness, through the comparison between these two metrics: results will be discussed directly in terms of taxonomic richness and evenness.
Both indices belong to the set of HCDT or generalized entropy, respectively corresponding to the 0 and 2 order of diversity (q), recommended for diversity studies [@Marcon2015]. 
The functional diversity was reported using the functional richness and functional evenness, *i.e* the Rao index of quadratic entropy.
The Rao index combines species abundance distribution, and the average pairwise functional dissimilarity between species computed by the gower distance.

@@

From 1984, the first pre-disturbance inventory, to 2015 (28 years after disturbance), 828 388 individual trees and 591 botanical species pertaining to 223 genera and 64 families were recorded.

@@

par(mfrow=c(2,2),mar=c(2,2.5,2.5,1),oma=c(1,2,1,4),no.readonly = T)
TaxoCompo(MatrepTaxo)
mtext("Taxonomic composition",side=3,line=1.8,cex=1.05)
FunCompo(MatrepFun)
mtext("Functional composition",side=3,line=1.8,cex=1.05)
EuclidDist(TaxoEuclid)
mtext("Euclidean distance\nfrom 1984 inventory",side=2,padj=0,line=2,cex=0.8)
mtext("(c)",side=3,adj=0,line=0.5)
EuclidDist(FunEuclid)
mtext("(d)",side=3,adj=0,line=0.5)
mtext("Years since disturbance",side=1,line=2.2,adj=1,cex=0.8)
legend("right",inset=-0.4,xpd=NA,legend=c("Control","Low","Interm.","High"),col=ColorsTr,lwd=2.5,bty="n",title="Treatment", cex=0.85)

@@

```{r DivTaxo, out.width = '100%', echo=FALSE, fig.cap="Trajectories of community taxonomic richness (ref:Bolda), Simpson diversity (ref:Boldb), functional richness (ref:Boldc), and Rao diversity (ref:Boldd). Values correspond to the difference over 30 years of community diversity with the values of reference 1984 pre-disturbance inventories. Shaded areas are the credibility intervals ", fig.env="figure*"}
load("DB/ReplacementTraj_ForGraphs")
load("DB/FunctionalTraj_ForGraphs")
load("DB/FunctionalRichnessTraj_ForGraphs")

@@

Our analysis highlighted the decoupling between functional and taxonomic  trajectories.
While initial differences in community taxonomic composition were maintained after disturbance, their trajectories in functional composition were similar.
In parallel, while the different taxonomic diversity trajectories displayed marked differences according to the disturbance intensity, the functional diversity trajectories were homogeneous whenever the disturbance intensity.
This decoupling was accompanied by a decrease in community functional redundancy which appeared to be the key element of community post-disturbance recovery.

@@

Before disturbance, the different communities had different taxonomic compositions, visualized by their distinct starting points on the NMDS axis 2.
These initial differences were maintained all along the 30 years following disturbance, with the disturbance leading a displacement on the NMDS axis 1.
The taxonomic compositional changes following disturbance were similar among communities, which may correspond to the recruitment of a group of pioneers, as *Cecropia* spp., *Vismia* spp., common to all communities whenever their initial differences and the intensity of disturbance [@Denslow2000;@Bongers2009].
Taxonomic trajectories continued with a recovery of the initial composition which, althought not fully achieved after 30 years, suggested the resilience of the taxonomic composition resilience and the maintenance of the initial differences among communities [@Folke2006].
The  return towards initial taxonomic composition showed that species not belonging to the pre-disturbance community were rarely recruited, probably because of the common dispersal limitations among tropical tree species [@Svenning2005].

@@

Comparing community diversity along time after disturbance revealed constrasting trajectories according to the distrubance intensity.
Below a disturbance intensity threshold, in the case of treatments 1 and 2, the community taxonomic richness increased.
In parallell, after a first increase, the taxonomic eveness stabilized or strated to decrease after 30 years.
Below the intensity threshold 
Contrastingly, above an intensity threshold, in the case of treatment 3, the taxonomic richness did not exceeded the initial value and the taxonomic evenness continiously increased for 30 years without sign of return towards the initial value.
Below and above a threshold, corresponding to a disturbance intensity between treatments 2 and 3, community taxonomic richness and evenness changed markedly.
A the disturbance intensity threshold, the taxonomic richness was maximized and the taxonomic evenness remained resilient. 
The disturbance intensity impacted the balance in the community between pre-disturbance surviving trees, and trees recruited afterward. 
On the one hand, below a 20%-25% AGB loss, the trees surviving after disturbance remained numerous enough to maintain the high taxonomic richness of the pre-disturbance community [@Bongers2009].
The recruitment of pioneers, infrequent or absent before disturbance, then increased the taxonomic richness all the more so that the disturbance was intense [@Martin2015; @Chaudhary2016].
As these pioneers became more dominant they balanced the usual hyper-dominance of tropical forests, hence temporarily increasing the taxonomic evenness [@Baraloto2012a].
On the other hand, above the intensity threshold, the taxonomic richness of surviving trees was too low to be offset by the recruitment of pioneers. 
In the Guiana shield indeed, the pool of true pioneers specifically recruited after disturbance is restricted to a few common genera (e.g. *Cecropia* spp., *Vismia* spp.) [@Guitet2018].
The taxonomic richness then decreased following intense disturbance, all the more so that the disturbance intensity was high [@Molino2001]. 
At the same time, pioneers became persistently dominant and prevented a return towards the initial evenness and the recovery of hyper-dominant shade-tolerant species.

@@

The decoupling between taxonomic and functional trajectories was accompanied by a decrease in the functional redundancy within the pre-disturbance functional space.
This decrease was explained by a loss of species following disturbance and then a lower redundancy of their functional characteristics.
This decrease of the functional redundancy was not compensated by the recruitment of pioneers, as those were functionally different from the pre-disturbance community.
Progressively though, more competitive,long-lived pioneers or late-successional species functionally closer to the pre-disturbance community progressively replaced  the first established species and restored the functional redundancy.
This replacement followed the lottery recruitment rules, implying an easy recruitment for the first species but becoming increasingly hampered by the emergence of interspecific competition [@Busing2002]. 
The recovery of the functional redundancy then relied upon the random process of species recruitment and was increasingly slow and difficult to anticipate [@Elmqvist2003;@Diaz2005].

@@

Community post-disturbance trajectories in compositon and diversity were shaped by the recruitment of a determined pool of pioneers identical among communities and disturbance intensity.
Nevertheless, taxonomic and functional composition trajectories appeared decoupled.
While functional trajectories remained similar in the functional space and converged towards the recovery of a comparable initial state, taxonomic trajectories showed initial differences in community composition that were maintained along time.
Community high functional redundancy could explain this decoupling as the loss of a species does not necessarily the loss of its functional charactristics in the community.
In this sense, the functional redundancy would mitigate the functional impact of disturbance.
Community functional and taxonomic diversity response to disturbance were contrasted as well.
While the functional trajectories remained similar whenever the disturbance intensity, taxonomic trajectories were markedly different above a threshold of 20-25% AGB removed for which taxonomic richness was maximized and taxonomic eveness remained resilient.
The Intermediate Disturbance Hypothesis applied to the linkage of disturbance with the taxonomic diversity, but not with the functional diversity.
Taxonomic diversity was maximized following a disturbance of intermediate intensity, that was the threshold of 25% AGB removed, and at an intermediate time after disturbance, that was around 25 years after disturbance when the recruitment of early- and late-successional species were balanced.
Whenever the disturbance intensity, community resilience (in terms of recovery of the pre-disturbance state) was tangible but required several decades and relied upon the random lottery recruitment of rare species. 
Although resilient, the functional redundancy was lower for more than 30 years after disturbance which cautioned against the risks of infrequent species loss and the persistence of disturbance-specific communities [@Herault2018].
Our conclusions first highlighted risks to lose rare species with visible impacts only in the long term, and led us to advice paying particular attention to rare and undangered species and to conduct regular inventories during exploitation times.
Second, post-disturbance trajectories seemed to greatly rely on post-disturbance recruitment dynamics that would be key to model post-disturbance trajectories.
